# Grouping and summarizing

## Sumarizing the median life expectancy

```{r}
gapminder %>%
  summarise(medianLifeExp = median(lifeExp))

# Filter for 1957 then summarize the median life expentancy
gapminder %>%
  filter(year == 1957) %>%
  summarise(medianLifeExp = median(lifeExp))

gapminder %>%
  filter(year == 1957) %>%
  summarise(medianLifeExp = median(lifeExp), maxGdpPerca = max(gdpPercap))
```

## Summarizing by year

**group_by** permite agrupar y luego aplicar resumen

```{r}
gapminder %>%
  group_by(year) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPerca = max(gdpPercap))
```

## Summarizing by continent

```{r}
gapminder %>%
  filter(year == 1957) %>%
  group_by(continent) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap)) %>%
  arrange(maxGdpPercap)
```

## Summarizing by continent and year

```{r}
gapminder %>%
  group_by(continent, year) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
```

# Visualizing summarazing

## Visualizing median life expectancy over time

```{r}
by_year <- gapminder %>%
  group_by(year) %>%
  summarise(medianGdpPercap = median(gdpPercap))

by_year

ggplot(by_year, aes(x = year, y=medianGdpPercap)) +
  geom_line()
```

```{r}
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarise(medianGdpPercap = median(gdpPercap))

ggplot(by_year_continent, aes(x=year, y=medianGdpPercap, color=continent)) +
  geom_line()
```

## Bar Plots

## Visualizing median GDP perr capita by continent

```{r}
by_continent <- gapminder %>%
  filter(year == 1952) %>%
  group_by(continent) %>%
  summarise(medianGdpPercap = median(gdpPercap))

by_continent

ggplot(by_continent, aes(x=continent, y=medianGdpPercap)) +
  geom_col()
```

## Visualizing GDP per capita by country in Oceania

```{r}
oceania_1952 <- gapminder %>%
  filter(continent == "Oceania", year == 1952)

oceania_1952

ggplot(oceania_1952, aes(x=country, y=gdpPercap)) +
  geom_col()
```

# Histogram

## Visualizing population

```{r}
library(gapminder)
library(ggplot2)
library(dplyr)

gapminder

gapminder_1952 <- gapminder %>%
  filter(year==1952) %>%
  mutate(pop_by_mil = pop/1000000)

gapminder_1952

ggplot(gapminder_1952, aes(x=pop_by_mil)) +
  geom_histogram(bins = 50)
```

## Visualizing population with x-axis on a log scale

```{r}
gapminder_1952 <- gapminder %>%
  filter(year==1952)

ggplot(gapminder_1952, aes(x=pop)) + 
  geom_histogram(bins=50) +
  scale_x_log10()
```

# Boxplots

## Comparing GDP per capita across continents

```{r}
gapminder_1952 <- gapminder %>%
  filter(year==1952)

ggplot(gapminder_1952, aes(x=continent, y=gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

## Adding a title to your graph

```{r}
ggplot(gapminder_1952, aes(x=continent, y=gdpPercap)) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Comparing GDP per capita across continents")
```

# Data manipulation with dplyr

```{r}
library(tidyverse)
counties <- read.csv("counties.csv")
glimpse(counties) # Panorama de los datos
counties
```

## Selecting columns

```{r}
counties %>%
  select(state, county, population, men, women)

counties_selected <- counties %>%
  select(state, county, population, private_work, public_work, self_employed)

counties_selected %>%
  arrange(desc(public_work))

counties_selected <- counties %>%
  select(state, county, population)

counties_selected %>%
  filter(population > 1000000)

counties_selected <- counties %>%
  select(state, county, population, private_work, public_work, self_employed)

counties_selected %>%
  filter(state=="Texas", population>10000) %>%
  arrange((desc(private_work)))
```

## Select, mutate, filter and arrange

```{r}
counties %>%
  select(state, county, population, men, women) %>%
  mutate(proportion_men = men/population) %>%
  filter(population > 10000) %>%
  arrange(desc(proportion_men))
```

# Aggregating Data

```{r}
select_counties <- counties %>%
  select(state, region, county, population, citizens)
```

## The verb count

```{r}
select_counties <- counties %>%
  select(state, region, county, population, citizens)

select_counties %>%
  count(region, sort = TRUE) %>%
  arrange(desc(n))
```

## Counting citizens by state

```{r}
counties_selected <- counties %>%
  select(state, region, county, population, citizens)

counties_selected %>%
  count(state, wt=citizens) %>% # Agrupa por state, pero suma valores de citizens
  arrange(desc(n))
```

## Mutating and counting

```{r}
counties_selected <- counties %>%
  select(state, region, county, population, walk)

counties_selected %>%
  mutate(population_walk = population*walk/100) %>%
  count(state, wt=population_walk, sort=TRUE)
```

## Summarizing

```{r}
counties_selected <- counties %>%
  select(county, population, income, unemployment)

counties_selected %>%
  summarize(min_population=min(population), max_unemployment=max(unemployment), average_income=mean(income))
```

## Summarizing by state

```{r}
counties_selected <- counties %>%
  select(county, population, state, land_area)

counties_selected %>%
  group_by(state) %>%
  summarise(total_area=sum(land_area), total_population=sum(population)) %>%
  mutate(density=total_population/total_area) %>%
  arrange(desc(density))

counties_selected <- counties %>%
  select(county, population, state, region)

counties_selected %>%
  group_by(region, state) %>%
  summarise(total_population=sum(population)) %>%
  summarise(average_population=mean(total_population), median_population=median(total_population))
```

## The top_n verb

```{r}
counties_selected <- counties %>%
  select(region, state, county, metro, population, walk)

counties_selected %>%
  group_by(region) %>%
  top_n(1, walk)

counties_selected <- counties %>%
  select(region, state, county, population, income)

counties_selected %>%
  group_by(region, state) %>%
  summarise(average_income=mean(income)) %>%
  top_n(1, average_income)
  
```

## The ungroup verb

```{r}
counties_selected <- counties %>%
  select(state, metro, population)

counties_selected %>%
  group_by(state, metro) %>%
  summarise(total_population=sum(population)) %>%
  top_n(1, total_population) %>%
  ungroup() %>%
  count(metro)
```

## Selecting columns

```{r}
counties %>%
  select(state, county, population, professional:production) %>%
  arrange(desc(service))

counties %>%
  select(state, county, population, ends_with("work")) %>%
  filter(public_work >= 50)

counties %>%
  select(state, county, population, starts_with("income"))

counties %>%
  select(state, county, population, contains("at"))
```

# Renaiming a column after count

```{r}
counties %>%
  count(state) %>%
  rename(num_conties = n) # Cambia el nombre de la columna

counties %>%
  select(state, county, poverty_rate = poverty)
```

# Using tansmute

```{r}
counties %>%
  transmute(state, county, population, density=population/land_area) %>%
  filter(population > 1000000) %>%
  arrange(desc(density))
```

# Chosing among the four verbs

```{r}
# Change the name of the unemployment column

counties %>%
  rename(unemployment_rate=unemployment)

# Keep the state and county columns, and the columns containnig poverty

counties %>%
  select(state, county, contains("poverty"))

# Calculate the franction woemen column without dropping the other columns

counties %>%
  mutate(fraction_women = women/population)

# Keep only the state, county, and emplyment_rate columns

counties %>%
  transmute(state, county, employment_rate = employed/population)
```
