# Grouping and summarizing

## Sumarizing the median life expectancy

```{r}
gapminder %>%
  summarise(medianLifeExp = median(lifeExp))

# Filter for 1957 then summarize the median life expentancy
gapminder %>%
  filter(year == 1957) %>%
  summarise(medianLifeExp = median(lifeExp))

gapminder %>%
  filter(year == 1957) %>%
  summarise(medianLifeExp = median(lifeExp), maxGdpPerca = max(gdpPercap))
```

## Summarizing by year

**group_by** permite agrupar y luego aplicar resumen

```{r}
gapminder %>%
  group_by(year) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPerca = max(gdpPercap))
```

## Summarizing by continent

```{r}
gapminder %>%
  filter(year == 1957) %>%
  group_by(continent) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap)) %>%
  arrange(maxGdpPercap)
```

## Summarizing by continent and year

```{r}
gapminder %>%
  group_by(continent, year) %>%
   summarise(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
```

# Visualizing summarazing

## Visualizing median life expectancy over time

```{r}
by_year <- gapminder %>%
  group_by(year) %>%
  summarise(medianGdpPercap = median(gdpPercap))

by_year

ggplot(by_year, aes(x = year, y=medianGdpPercap)) +
  geom_line()
```

```{r}
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarise(medianGdpPercap = median(gdpPercap))

ggplot(by_year_continent, aes(x=year, y=medianGdpPercap, color=continent)) +
  geom_line()
```

## Bar Plots

## Visualizing median GDP perr capita by continent

```{r}
by_continent <- gapminder %>%
  filter(year == 1952) %>%
  group_by(continent) %>%
  summarise(medianGdpPercap = median(gdpPercap))

by_continent

ggplot(by_continent, aes(x=continent, y=medianGdpPercap)) +
  geom_col()
```

## Visualizing GDP per capita by country in Oceania

```{r}
oceania_1952 <- gapminder %>%
  filter(continent == "Oceania", year == 1952)

oceania_1952

ggplot(oceania_1952, aes(x=country, y=gdpPercap)) +
  geom_col()
```
